---
title: "Exercises Day 2"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data

Read in the gapminder dataset

```{r}
setwd("..")
gapminder <- read_csv("data/gapminder5.csv")
```

## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

```{r}
year <- 2002
if (any(gapminder$year == year)){
    print(paste("There are records from", year))
} else {
    print(paste("There are no records from", year))
}

# For 2012, set the year <- 2012
```

## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r}
countries <- unique(gapminder$country)

for (i in countries){
    if (gapminder$pop[gapminder$country == i] < mean(gapminder$pop)){
        mean_le <- mean(gapminder$lifeExp[gapminder$country == i])
        print(paste("Country", i, ":", mean_le))
    }
}
```

Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70

```{r}

```

## Including Plots

You can also embed plots, for example:

```{r another chunk name}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
