---
title: "R Day 2 exercises"
author: "Catherine Yang"
date: "`r Sys.Date()`"
output: html_document
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Read both California energy datasets. Make sure the `datetime` variable is in an appropriate data type (i.e. not character). 

```{r}
setwd("..")
library(lubridate)
library(readr)

# Load data
generation <- read_csv("data/ca_energy_generation.csv")
imports <- read_csv("data/ca_energy_imports.csv")
```


## Merge and reshape the data

Merge the two datasets and then melt the resulting dataframe/datatable to make it tidy.

```{r}
# Merge the generation and imports dataset
merged_energy <- merge(generation, imports, by="datetime") 

# Melt the merged dataset to long form
merged_energy <- melt(merged_energy, id.vars="datetime",
         variable.name="source",
         value.name="usage")
```


## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Bonus: If you are using `dplyr`, try to do this all in one pipe!

```{r}
# Add the day and log_output variables
merged_energy <- mutate(merged_energy, day=as_date(datetime), log_usage=log(usage))

# Challenge question
merged_energy %>% 
    group_by(day) %>% 
    mutate(total_daily_output=sum(usage, na.rm=TRUE)) %>% 
    ungroup() %>% 
    mutate(per_output=usage/total_daily_output)
```


## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the `dplyr` verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order. The `data.table` function is `setorder`.) Which has the least?
2. Which source has the greatest mean output by day? Which has the least? (Do not include zero values.)
3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least? (Do not include zero values.)

```{r}
# Question 1
merged_energy %>% 
    group_by(source) %>%
    summarize(mean_usage=mean(usage, na.rm=TRUE)) %>%
    arrange(desc(mean_usage))
```

Natural gas has the greatest mean output by hour; coal has the least

```{r}
# Question 2
merged_energy %>%
    group_by(source, day) %>%
    summarize(mean_usage=mean(usage, na.rm=TRUE)) %>%
    arrange(desc(mean_usage))
```

Natural gas has the greatet mean usage by day; coal has the least

```{r}
# Question 3
merged_energy %>%
    group_by(source) %>% 
    summarize(var_usage=var(usage, na.rm=TRUE)) %>% 
    arrange(desc(var_usage))
```

Natural gas has the greatest variance; coal has the least

## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with yourdata manipulation skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by?

```{r}
# Load data
setwd("..")
regroup <- read_csv("data/ca_energy_regroup.csv")

# Merge the regroup dataframe with the merged_energy dataframe on regroup$type = merged_energy$source
energy <- merge(merged_energy, regroup, by.x="source", by.y="type")

# Check that the merge was as intended
head(energy)
```

```{r}
# Look at the total usage for the renewability types
energy %>% 
    group_by(group) %>% 
    summarize(total_usage=sum(usage, na.rm=TRUE)) %>% 
    arrange(desc(total_usage))

# Look at mean and variance of usage for the renewability types
energy %>% 
    group_by(group) %>% 
    summarize(mean_usage=mean(usage, na.rm=TRUE), var_usage=var(usage, na.rm=TRUE)) %>% 
    arrange(desc(mean_usage))
```
It looks like thermal and renewable energy types see the highest total usage, but on an average basis, imported energy has the highest average usage and renewable energy has the lowest average usage. This means that while thermal and renewable energy are used the most, imported energy sees the most usage per observation.

Nuclear energy has the lowest usage variance out of all energy types, which could mean that nuclear energy is less substitutable than the other energy types.

```{r}
# Look at the average energy usage for each group by hour to see if there's any trend in types of energy used throughout the day
energy %>%
    mutate(hour=hour(datetime)) %>%
    group_by(group, hour) %>%
    summarize(usage_by_hour=mean(usage, na.rm=TRUE)) %>%
    arrange(group, hour)
```


A few observations:
- Renewable energy usage picks up drastically around 8am to 5pm, which is during normal business hours. Perhaps commercial activities utilize renewable energy more heavily.
- Imported energy usage decreases from around 8am to 5pm, then picks back up starting around 6pm, which indicates that at home energy usage (outside of working hours), tends to utilize imported energy the most. This means that there is opportunity for residents to utilize more natural and/or renewable energy sources for at home usage, instead of imported energy which can be costly and environmentally unfriendly.
- Thermal energy usage shows increasing usage from ~10am to 6pm, which is during the hours where temperature is probably the highest.



